|

### 2021-09-22 publication/manuscript.Rmd
move encoder algos and dataset table to the supplementary material

Pseudocode is presented in the main text (for non-standard encodings), or the Supplementary Material (for standard methods).
An important detail is how encoding techniques treat new levels during the prediction phase.

@@

\begin{algorithm}
  \caption{Leaf Encoding}
  \label{alg:leaf}
  \begin{algorithmic}
    \State \underline{Training:} require number of cross-validation folds $K \in \mathbb{N}$
    \State fit CART tree on $\mathcal{D}^{train}$ with complexity pruning based on $K$-fold cross-validation
      \ForAll{$x^{train}_i \in \boldsymbol{x}^{train}$}
        $\tilde{x}_i = t$ with $x^{train}_i$ in terminal node $t$
      \EndFor
    \State \underline{Prediction:}
    \For{$x^{new}$}
        \If{$x^{new} \in \mathcal{L}^{train}$}\ {$\tilde{x}^{new} = t$ with $x^{new}$ in terminal node $t$}
        \Else\ {$\tilde{x}^{new} = b$ where $b$ indicates the biggest terminal node}
        \EndIf
    \EndFor
  \end{algorithmic}
\end{algorithm}

@@

Smoothed target encoding [@micci_barreca_2001] can be interpreted as a simple (generalized) linear mixed model (glmm) in which the target is predicted by a random intercept for each feature level in addition to a fixed global intercept.
This connection is described in @kuhn_2019.
To achieve regularized impact encoding, we implemented glmm encoders for regression, binary, and multiclass classification which are described in Algorithm \ref{alg:glmm-regr} (regression) and in the Supplementary Material (classification).
The encoded value for each level is based on the spherical conditional mode estimates.
In regression, the conditional modes are similar to the mean target value for each level, weighted by the relative observed frequency of that level in the training set [@gelman_2006, @bates_2018].
The estimate of the fixed intercept can be used during the prediction phase to encode new feature levels not observed in the training set.
In multiclass classification, we fit $C$ one vs. rest glmms resulting in one encoded feature per class.
An important advantage of using a glmm over impact encoding with a smoothing parameter is that a reasonable amount of regularization is determined automatically and tuning the complete ML pipeline is not necessary.

Overfitting can be further avoided through combination with cross-validation (CV) to train the encoder on independent observations without limiting the data to train the ML model.
We provide an implementation which combines target encoding based on glmms with CV.
During the training phase, we partition the data using CV into $n.folds$ and fit a glmm on each resulting training set.
For each observation, there is exactly one glmm that did not use that observation for model fitting and can be safely used for encoding.
Note that the $n.folds$ CV models (for $n.folds$ > 1) are only used during the training phase.
In the prediction phase, feature values are always encoded by a single glmm fitted to the complete training set.
We study this method in three different settings: without CV (noCV), with $5-$ (5CV) and with $10-$ fold CV (10CV).
In our study, we use the lmer (regression) and glmer (classification) functions from the lme4 package in R [@bates_2015] as an efficient way to fit glmms.
\begin{algorithm}
  \caption{GLMM Encoding Regression}
  \label{alg:glmm-regr}
  \begin{algorithmic}
    \State \underline{Training:} require $n.folds \in \mathbb{N}$
     \State fit simple random intercept model: $y^{train}_i = \beta_{0l} + \epsilon_{i} = \gamma_{oo} + u_l + \epsilon_{i}$ on $\mathcal{D}^{train}$
     \State with $u_l \overset{iid}{\sim} N(0,\tau^2)$, $\epsilon_i \overset{iid}{\sim} N(0,\sigma^2)$ and $x^{train}_i = l$, $l \in \mathcal{L}^{train}$
    \If{$n.folds = 1$}
      \ForAll{$x^{train}_i \in \boldsymbol{x}^{train}$}
        \State $\hat{x}^{train}_i = \hat{\beta}_{0l}^{\mathcal{D}^{train}}$ with $x^{train}_i = l$
      \EndFor
    \Else\ {use $n.folds$ cross-validation scheme to make training sets $\mathcal{D}^{train}_1, \dots, \mathcal{D}^{train}_{n.folds}$}
      \State and fit simple random intercept model on each $\mathcal{D}^{train}_{m}$
      \ForAll{$x^{train}_i \in \boldsymbol{x}^{train}$}
        \State $\hat{x}^{train}_i = \hat{\beta}_{0l}^{\mathcal{D}^{train}}$ with $x^{train}_i = l$ based on the model m
        \State with $(x^{train}_i, y^{train}_i) \notin \mathcal{D}^{train}_m$
      \EndFor
    \EndIf
    \State \underline{Prediction:}
    \For{$x^{new}$}
      \If{$x^{new} \in \mathcal{L}^{train}$}
        \State $\hat{x}^{new} = \hat{\beta}_{0l}^{\mathcal{D}^{train}}$ with $x^{new} = l$ based on full model fitted on $\mathcal{D}^{train}$
      \Else\ {$\hat{x}^{new} = \hat{\gamma}_{00}$ based on full model fitted on $\mathcal{D}^{train}$}
      \EndIf
    \EndFor
  \end{algorithmic}
\end{algorithm}

@@

Table \ref{tab:datasets} summarizes datasets used in our benchmark.
We specifically investigate datasets that contain categorical variables with a large number of levels, including many well-known datasets
from previous studies [@cerda_2018; @prokhorenkova_2018; @coors_2018; @kuhn_2019].
All datasets can be downloaded from the `OpenML` platform [@vanschoren_2013] based on the name or the displayed OmlId.
The datasets include `r nrow(descr_dat[Classes == 0])` regression, `r nrow(descr_dat[Classes == 2])` binary classification, and `r nrow(descr_dat[Classes > 2])` multiclass classification problems (between `r descr_dat[Classes > 2][, min(Classes)]` and `r descr_dat[Classes > 2][, max(Classes)]` classes).
To assess the imbalance in categorical variables we computed the normalized entropy for each categorical variable.
The maximum normalized entropy is $1$, which corresponds to a uniform distribution, while a lower number indicates a larger imbalance.
Sample sizes range between `r descr_dat[, min(Obs)]` and `r descr_dat[, max(Obs)]` observations.
The total number of features ranges between `r descr_dat[, min(NumFeats + BinFeats + CatFeats)]` and `r descr_dat[, max(NumFeats + BinFeats + CatFeats)]`.
Datasets contain between `r descr_dat[, min(sapply(strsplit(HighCardLevels, " "), length))]` and `r descr_dat[, max(sapply(strsplit(HighCardLevels, " "), length))]` categorical features with more than $10$ levels, with the maximum number of levels for a feature ranging between `r descr_dat[, min(MaxCatLvls)]` and `r descr_dat[, max(MaxCatLvls)]`. Missing values are present in about half of the datasets.

@@

```{r datasets, warning=FALSE, results="asis", fig.pos="h"}
library(kableExtra)
landscape(
  knitr::kable(
    descr_dat2[, .(OmlId, Name, Cl = Classes, N = Obs, `NA%` = NAs*100/(Obs*(NumFeats + BinFeats + CatFeats)),
                Num = as.integer(NumFeats), Bin = as.integer(BinFeats), Cat = as.integer(CatFeats), HighCardLevels, Entropy = "")],
    digits=2L,
    caption="Benchmark Datasets and Dataset Characteristics",
    label="datasets"
  ) %>%
  column_spec(10, image = spec_boxplot(lapply(descr_dat2$entropies, unlist))) %>%
  footnote(general = "OmlId = Id on OpenML, Name = name on OpenML, Cl = classes (0: regression), N = observations, NA% = percentage of missing values, Num = numeric features, Bin = binary features, Cat = categorical features, HighCardLevels = number of levels for each categorical feature with at least 10 levels (some levels are not displayed for KDDCup09_upselling), Entropy = box-plot of normalized Shannon-entropy across levels (smaller = larger imbalance).", threeparttable = TRUE) %>%
  kable_styling(latex_options="scale_down")
)
```

@@

(ref:rank-res) Consensus rankings across all datasets for each algorithm. Lower ranks indicate better performance. The rank of the *none* control condition of the RF (rank 11) was omitted from the figure.

>>>>

Pseudocode for all encoders is presented in the Supplementary Material.
An important detail is how encoding techniques treat new levels during the prediction phase.

@@

@@

Smoothed target encoding [@micci_barreca_2001] can be interpreted as a simple (generalized) linear mixed model (glmm) in which the target is predicted by a random intercept for each feature level in addition to a fixed global intercept.
This connection is described in @kuhn_2019.
To achieve regularized impact encoding, we implemented glmm encoders for regression, binary, and multiclass classification.
The encoded value for each level is based on the spherical conditional mode estimates.
In regression, the conditional modes are similar to the mean target value for each level, weighted by the relative observed frequency of that level in the training set [@gelman_2006, @bates_2018].
The estimate of the fixed intercept can be used during the prediction phase to encode new feature levels not observed in the training set.
In multiclass classification, we fit $C$ one vs. rest glmms resulting in one encoded feature per class.
An important advantage of using a glmm over impact encoding with a smoothing parameter is that a reasonable amount of regularization is determined automatically and tuning the complete ML pipeline is not necessary.

Overfitting can be further avoided through combination with cross-validation (CV) to train the encoder on independent observations without limiting the data to train the ML model.
We provide an implementation which combines target encoding based on glmms with CV.
During the training phase, we partition the data using CV into $n.folds$ and fit a glmm on each resulting training set.
For each observation, there is exactly one glmm that did not use that observation for model fitting and can be safely used for encoding.
Note that the $n.folds$ CV models (for $n.folds$ > 1) are only used during the training phase.
In the prediction phase, feature values are always encoded by a single glmm fitted to the complete training set.
We study this method in three different settings: without CV (noCV), with $5-$ (5CV) and with $10-$ fold CV (10CV).
In our study, we use the lmer (regression) and glmer (classification) functions from the lme4 package in R [@bates_2015] as an efficient way to fit glmms.

@@

A table showing a detailed summary of all benchmark datasets can be found in the Supplementary Material.
We specifically investigate datasets that contain categorical variables with a large number of levels, including many well-known datasets
from previous studies [@cerda_2018; @prokhorenkova_2018; @coors_2018; @kuhn_2019].
All datasets can be downloaded from the `OpenML` platform [@vanschoren_2013] based on the name or the displayed OmlId.
The datasets include `r nrow(descr_dat[Classes == 0])` regression, `r nrow(descr_dat[Classes == 2])` binary classification, and `r nrow(descr_dat[Classes > 2])` multiclass classification problems (between `r descr_dat[Classes > 2][, min(Classes)]` and `r descr_dat[Classes > 2][, max(Classes)]` classes).
To assess the imbalance in categorical variables we computed the normalized entropy for each categorical variable.
The maximum normalized entropy is $1$, which corresponds to a uniform distribution, while a lower number indicates a larger imbalance.
Sample sizes range between `r descr_dat[, min(Obs)]` and `r descr_dat[, max(Obs)]` observations.
The total number of features ranges between `r descr_dat[, min(NumFeats + BinFeats + CatFeats)]` and `r descr_dat[, max(NumFeats + BinFeats + CatFeats)]`.
Datasets contain between `r descr_dat[, min(sapply(strsplit(HighCardLevels, " "), length))]` and `r descr_dat[, max(sapply(strsplit(HighCardLevels, " "), length))]` categorical features with more than $10$ levels, with the maximum number of levels for a feature ranging between `r descr_dat[, min(MaxCatLvls)]` and `r descr_dat[, max(MaxCatLvls)]`. Missing values are present in about half of the datasets.

@@

```{r datasets, warning=FALSE, results="hide", fig.pos="h"}
library(kableExtra)
landscape(
  knitr::kable(
    descr_dat2[, .(OmlId, Name, Cl = Classes, N = Obs, `NA%` = NAs*100/(Obs*(NumFeats + BinFeats + CatFeats)),
                Num = as.integer(NumFeats), Bin = as.integer(BinFeats), Cat = as.integer(CatFeats), HighCardLevels, Entropy = "")],
    digits=2L,
    caption="Benchmark Datasets and Dataset Characteristics",
    label="datasets"
  ) %>%
  column_spec(10, image = spec_boxplot(lapply(descr_dat2$entropies, unlist))) %>%
  footnote(general = "OmlId = Id on OpenML, Name = name on OpenML, Cl = classes (0: regression), N = observations, NA% = percentage of missing values, Num = numeric features, Bin = binary features, Cat = categorical features, HighCardLevels = number of levels for each categorical feature with at least 10 levels (some levels are not displayed for KDDCup09_upselling), Entropy = box-plot of normalized Shannon-entropy across levels (smaller = larger imbalance).", threeparttable = TRUE) %>%
  kable_styling(latex_options="scale_down")
)
```

@@

(ref:rank-res) Consensus rankings across all datasets for each algorithm. Lower ranks indicate better performance. The rank of the *none* control condition of the RF (rank 11 of 12) was omitted from the figure.

### 2021-09-22 publication/supplementary_material.Rmd
move encoder algos and dataset table to the supplementary material

@@

From the results shown in Table 2 we can observe, that for most algorithms and datasets, one-hot encoding is to be preferred over dummy-encoding.

>>>>

# Benchmark Datasets

```{r, echo=FALSE, results='hide', message=FALSE}
library(data.table)
library(stringi)
library(kableExtra)
descr_dat = as.data.table(readRDS("../analysis/high_cardinality_benchmark/descr_dat.rds"))
# remove unfinished datasets
descr_dat = descr_dat[!(Name %in% c("KDD98", "Traffic_violations", "sf-police-incidents")),]
```

```{r, eval=FALSE, echo=FALSE}
# Collect info (do not run, just here for reproducibility)
if (FALSE) {
  # Collect cardinality information
  library(OpenML)
  setOMLConfig(arff.reader = "farff")
  out = list()
  for (id in descr_dat$OmlId) {
    d = getOMLDataSet(id)
    fcts = (sapply(d$data, class) == "factor") & (names(d$data) != d$desc$default.target.attribute)
    out[[as.character(id)]] = lapply(d$data[fcts], table)
    saveRDS(out, "../analysis/high_cardinality_benchmark/dataset_factor_levels.rds")
    gc()
  }

  # Restart
  setOMLConfig(arff.reader = "RWeka") # farff seems to break for one dataset
  out = readRDS("../analysis/high_cardinality_benchmark/dataset_factor_levels.rds")
  for (id in sample(setdiff(descr_dat$OmlId, as.numeric(names(out))))) {
    d = getOMLDataSet(id)
    fcts = (sapply(d$data, class) == "factor") & (names(d$data) != d$desc$default.target.attribute)
    out[[as.character(id)]] = lapply(d$data[fcts], table)
    saveRDS(out, "../analysis/high_cardinality_benchmark/dataset_factor_levels.rds")
    gc()
  }
}
```

```{r, echo=FALSE}
# Normalized Shannon entropy
shannonEntropy <- function(counts.table, log.base = getOption("GeneFamilies.entropy.log.base",
    base::exp(1))) {
    if (length(counts.table) <= 1)
        return(0)
    c.t.s <- sum(counts.table)
    -sum(sapply(counts.table, function(x) x/c.t.s * log(x/c.t.s, base = log.base)))/log(length(counts.table),
        base = log.base)
}
out = readRDS("../analysis/high_cardinality_benchmark/dataset_factor_levels.rds")
ents = lapply(out, function(x) lapply(x, shannonEntropy))
ent2 = lapply(out, function(x) lapply(x, function(x) min(x)/max(x)))
ents = data.table(OmlId = as.numeric(names(ents)), entropies = ents)
descr_dat = merge(descr_dat, ents)
```

```{r, echo=FALSE}
descr_dat2 = descr_dat
descr_dat2[Classes == 0, task_type := rep("regr", .N)]
descr_dat2[Classes == 2, task_type := rep("bin_class", .N)]
descr_dat2[Classes > 2, task_type  := rep("multi_class", .N)]
descr_dat2[, task_type := factor(task_type, levels = c("regr", "bin_class", "multi_class"))]
setorder(descr_dat2, task_type, Obs)
descr_dat2$HighCardLevels = stri_replace_all_fixed(descr_dat2$HighCardLevels, " ", "")
descr_dat2[Name == "KDDCup09_upselling", HighCardLevels := "14,...,5073,5713,13990,15415,15415"]
#hcl_list = lapply(descr_dat$HighCardLevels, function(x) {log(as.numeric(unlist(strsplit(x, #split = ","))))})
#descr_dat2$logNHCL = ""
```

```{r datasets, echo=FALSE, warning=FALSE, results="asis", fig.pos="h"}
library(kableExtra)
landscape(
  knitr::kable(
    descr_dat2[, .(OmlId, Name, Cl = Classes, N = Obs, `NA%` = NAs*100/(Obs*(NumFeats + BinFeats + CatFeats)),
                Num = as.integer(NumFeats), Bin = as.integer(BinFeats), Cat = as.integer(CatFeats), HighCardLevels, Entropy = "")],
    digits=2L,
    caption="Benchmark Datasets and Dataset Characteristics",
    label="datasets"
  ) %>%
  column_spec(10, image = spec_boxplot(lapply(descr_dat2$entropies, unlist))) %>%
  footnote(general = "OmlId = Id on OpenML, Name = name on OpenML, Cl = classes (0: regression), N = observations, NA% = percentage of missing values, Num = numeric features, Bin = binary features, Cat = categorical features, HighCardLevels = number of levels for each categorical feature with at least 10 levels (some levels are not displayed for KDDCup09_upselling), Entropy = box-plot of normalized Shannon-entropy across levels (smaller = larger imbalance).", threeparttable = TRUE) %>%
  kable_styling(latex_options="scale_down")
)
```

@@

From the results shown in Table \ref{tab:oneh-win} we can observe, that for most algorithms and datasets, one-hot encoding is to be preferred over dummy-encoding.
